Optimal (Multiway) Spatial Joins

### 2024-09-19 msj.tex
yufei

\begin{abstract}
    In a {\em spatial join}, we are given a constant number $k \geq 2$ of sets --- denoted as $R_1, R_2, ..., R_k$ --- containing axis-parallel rectangles in a 2D space. The objective is to report all $k$-tuples $(r_1, r_2, ..., r_k) \in R_1 \times R_2 \times ... \times R_k$ where the rectangles $r_1, r_2, ..., r_k$ have a non-empty intersection, i.e., $r_1 \cap r_2 \cap ... \cap r_k \neq \emptyset$. The problem holds significant importance in spatial databases and has been extensively studied in the database community. In this paper, we show how to settle the problem in $O(n \log n + \out)$ time --- regardless of the constant $k$ --- where $n = \sum_{i=1}^k |R_i|$ and $\out$ is the result size (i.e., the total number of $k$-tuples reported). The runtime  is asymptotically optimal in the class of comparison-based algorithms, to which our solution belongs. Previously, the state of the art was an algorithm with running time $O(n \log^{\red{2k-1}} n + \out)$.
\end{abstract}

@@

\extraspacing {\bf Math Conventions.} For any integer $x \ge 1$, we use $[x]$ to represent the set $\set{1, 2, ..., x}$. Given $k \ge 2$ sets $S_1, S_2, ..., S_k$ (of arbitrary elements), we often treat a $k$-tuple $(e_1, e_2, .., e_k)$ in the \red{Cartesian} product $S_1 \times S_2 \times ... \times S_k$ as a $k$-dimensional vector $\bm{t}$ with $\bm{t}[i] = e_i$ for each $i \in [k]$. Unless otherwise stated, every mention of the word ``rectangle'' henceforth will refer to an axis-parallel rectangle in $\real^2$. All logarithms have base 2 by default.

@@

In \cite{kcko22}, Khamis et al.\ investigated a type of joins that extends the conventional equi-join in two ways. First, each attribute value in a relation is an interval (rather than a real value); second, each equality predicate in an equi-join is replaced with a ``non-empty intersection'' predicate on the attributes involved. The $k$-SJ problem can be converted to a join under the framework of \cite{kcko22} as defined next. For each $i \in [k]$, define $R_i$ as a relation over two attributes $X$ and $Y$. For each tuple $\bm{t} \in R_i$, its values $\bm{t}(X)$ and $\bm{t}(Y)$ on the two attributes are both intervals (effectively defining a rectangle). The objective is to output all $k$-tuples $(\bm{t}_1, \bm{t}_2, ..., \bm{t}_k) \in R_1 \times R_2 \times ... \times R_k$ satisfying $\bigcap_{i=1}^k \bm{t}_i(X) \ne \emptyset$ and $\bigcap_{i=1}^k \bm{t}_i(Y) \ne \emptyset$. It is clear that there is one-one correspondence between the result of this join and that of k-SJ. Khamis et al.\ \cite{kcko22} developed an algorithm that can process the join  in $O(n \log^{\red{2k-1}} n + \out)$ time.

@@

\begin{table}
    \begin{tabular}{c|c|c|c}
        $\bm{k}$ & {\bf method} & {\bf runtime} & {\bf remark} \\
        \hline\hline
        2 & folklore & $O(n \log n + \out)$ & optimal \\
        \hline
        $\ge 3$ & before 2022 & $O(n^k)$ &  \\
        $\ge 3$ & \cite{ty22} & $O((n + \out) \cdot \polylog n)$ & \\
        $\ge 3$ & \cite{kcko22} & $O(n \log^{\red{2k-1}} n + \out)$ & \\
        \hline
        $\ge 3$ & ours & $O(n \log n + \out)$ & optimal
    \end{tabular}

@@

Given a rectangle $r = [x_1, x_2] \times [y_1, y_2]$, we define $\xleft(r) = x_1$, $\xright(r) = x_2$, $\ybot(r) = y_1$, and $\ytop(r) = y_2$. Consider a $k$-tuple $\bm{t} = (r_1, r_2, ..., r_k)$ where $k \ge 2$ and each $\bm{t}[i] = r_i$ ($i \le [k]$) is a rectangle. We define
\myeqn{
    B_\bm{t} &=& \bigcap_{i=1}^{\red{k}} r_i; \label{eqn:B_t}
}
namely, $B_\bm{t}$ is the intersection of the rectangles in $\bm{t}$ (note: $B_\bm{t}$ is a rectangle itself). Also, if $B_\bm{t}$ is not empty, define:
\myitems{
    \item $\gleft(\bm{t})$ as the rectangle $r_i$, $i \in [k]$, satisfying $\xleft(r_i)$ $= \xleft(B_\bm{t})$. In case multiple values in $[k]$ fulfill the condition, let $i$ be the smallest of such values.

@@

In the {\em reporting version} of Problem $\mathscr{A}$, the goal is to output, for each point $p \in P$, all the rectangles $r \in R$ containing $p$; if no such $r$ exists, report nothing for $p$. In Figure~\ref{fig:guard}a, for instance, the output is $\set{(p_2: r_1, r_2), (p_3: r_2)}$. As shown in Appendix~\ref{app:bricks}, \red{the} problem can be solved in $O(n \log n + \out)$ time, where $\out$ is the number of pairs $(p, r) \in P \times R$ such that $p \in r$.

@@

\begin{figure*}
    \begin{tabular}{cc}
        \includegraphics[height=30mm]{./artwork/prob-a} &
        \includegraphics[height=27mm]{./artwork/prob-b}  \\[2mm]
        (a) \red{Problem} $\mathscr{A}$ &
        \hspace{3mm}
        (b) Problem $\mathscr{B}$ \\[3mm]
    \end{tabular}
    %\vspace{3mm}
    \begin{tabular}{ccc}
        \includegraphics[height=30mm]{./artwork/prob-c} &
        \includegraphics[height=30mm]{./artwork/prob-d} &
        \includegraphics[height=30mm]{./artwork/prob-e} \\[2mm]
        (c) Problem $\mathscr{C}$ &
        (d) Problem $\mathscr{D}$ &
        (e) Problem $\mathscr{E}$
    \end{tabular}

@@

\extraspacing {\bf Problem $\bm{\mathscr{E}}$.} The input involves a set $H$ of horizontal segments and a set $R$ of rectangles. The goal is to report, for each segment $h \in H$, the set $\contain_R(h)$ --- defined in \eqref{eqn:contained} --- where the rectangles are sorted by their right boundaries; if $\contain_R(h)$ is empty, output nothing for $h$. Formally, if $r_1, r_2, ..., r_z$ for some $z \ge 1$ are all the rectangles in $\contain_R(h)$, we output $(h: r_1, r_2, ..., r_z)$, provided that $\xright(r_i) \ge \xright(r_{i-1})$ for each $i \in [2, z]$. Figure~\ref{fig:probs}e gives an example where $H = \set{h_1, h_2, h_3}$ and $R = \set{r_1, r_2, r_3}$; the output is $\set{(h_1: r_2, r_1), (h_2: \red{r_3, r_2})}$. In Appendix~\ref{app:bricks}, we explain how to solve the problem in $O(n \log n + \out)$ time where $n = |H| + |R|$ and $\out$ is the number of pairs $(h, r) \in H \times R$ such that $r$ contains $h$.

@@

\extraspacing {\underline{\em Example 4.1.}} To illustrate our algorithm, we will utilize the running example in Figure~\ref{fig:hv:type1:ex}a, where $k = 4$, and $R_1 = \set{\alpha}$ (the solid rectangle), $R_2 = \set{\red{\beta_1}, \beta_2}$ (the dashed rectangles), $H$ $=$ $\set{h_1,$ $h_2,$ $..., h_6}$, and $V = \set{v_1, v_2, ..., v_5}$. The set $\J_1$ contains the following tuples: $(\alpha, \beta_2, h_2, v_3)$, $(\alpha, \beta_2, h_2, v_5)$, $(\alpha, \beta_2, h_5, v_3)$, and $(\alpha,$ $\beta_2, h_5, v_5)$. \done

>>>>

\begin{abstract}
    In a {\em spatial join}, we are given a constant number $k \geq 2$ of sets --- denoted as $R_1, R_2, ..., R_k$ --- containing axis-parallel rectangles in a 2D space. The objective is to report all $k$-tuples $(r_1, r_2, ..., r_k) \in R_1 \times R_2 \times ... \times R_k$ where the rectangles $r_1, r_2, ..., r_k$ have a non-empty intersection, i.e., $r_1 \cap r_2 \cap ... \cap r_k \neq \emptyset$. The problem holds significant importance in spatial databases and has been extensively studied in the database community. In this paper, we show how to settle the problem in $O(n \log n + \out)$ time --- regardless of the constant $k$ --- where $n = \sum_{i=1}^k |R_i|$ and $\out$ is the result size (i.e., the total number of $k$-tuples reported). The runtime  is asymptotically optimal in the class of comparison-based algorithms, to which our solution belongs. Previously, the state of the art was an algorithm with running time $O(n \log^{2k-1} n + \out)$.
\end{abstract}

@@

\extraspacing {\bf Math Conventions.} For any integer $x \ge 1$, we use $[x]$ to represent the set $\set{1, 2, ..., x}$. Given $k \ge 2$ sets $S_1, S_2, ..., S_k$ (of arbitrary elements), we often treat a $k$-tuple $(e_1, e_2, .., e_k)$ in the Cartesian product $S_1 \times S_2 \times ... \times S_k$ as a $k$-dimensional vector $\bm{t}$ with $\bm{t}[i] = e_i$ for each $i \in [k]$. Unless otherwise stated, every mention of the word ``rectangle'' henceforth will refer to an axis-parallel rectangle in $\real^2$. All logarithms have base 2 by default.

@@

In \cite{kcko22}, Khamis et al.\ investigated a type of joins that extends the conventional equi-join in two ways. First, each attribute value in a relation is an interval (rather than a real value); second, each equality predicate in an equi-join is replaced with a ``non-empty intersection'' predicate on the attributes involved. The $k$-SJ problem can be converted to a join under the framework of \cite{kcko22} as defined next. For each $i \in [k]$, define $R_i$ as a relation over two attributes $X$ and $Y$. For each tuple $\bm{t} \in R_i$, its values $\bm{t}(X)$ and $\bm{t}(Y)$ on the two attributes are both intervals (effectively defining a rectangle). The objective is to output all $k$-tuples $(\bm{t}_1, \bm{t}_2, ..., \bm{t}_k) \in R_1 \times R_2 \times ... \times R_k$ satisfying $\bigcap_{i=1}^k \bm{t}_i(X) \ne \emptyset$ and $\bigcap_{i=1}^k \bm{t}_i(Y) \ne \emptyset$. It is clear that there is one-one correspondence between the result of this join and that of k-SJ. Khamis et al.\ \cite{kcko22} developed an algorithm that can process the join  in $O(n \log^{2k-1} n + \out)$ time.

@@

\begin{table}
    \begin{tabular}{c|c|c|c}
        $\bm{k}$ & {\bf method} & {\bf runtime} & {\bf remark} \\
        \hline\hline
        2 & folklore & $O(n \log n + \out)$ & optimal \\
        \hline
        $\ge 3$ & before 2022 & $O(n^k)$ &  \\
        $\ge 3$ & \cite{ty22} & $O((n + \out) \cdot \polylog n)$ & \\
        $\ge 3$ & \cite{kcko22} & $O(n \log^{2k-1} n + \out)$ & \\
        \hline
        $\ge 3$ & ours & $O(n \log n + \out)$ & optimal
    \end{tabular}

@@

Given a rectangle $r = [x_1, x_2] \times [y_1, y_2]$, we define $\xleft(r) = x_1$, $\xright(r) = x_2$, $\ybot(r) = y_1$, and $\ytop(r) = y_2$. Consider a $k$-tuple $\bm{t} = (r_1, r_2, ..., r_k)$ where $k \ge 2$ and each $\bm{t}[i] = r_i$ ($i \le [k]$) is a rectangle. We define
\myeqn{
    B_\bm{t} &=& \bigcap_{i=1}^{k} r_i; \label{eqn:B_t}
}
namely, $B_\bm{t}$ is the intersection of the rectangles in $\bm{t}$ (note: $B_\bm{t}$ is a rectangle itself). Also, if $B_\bm{t}$ is not empty, define:
\myitems{
    \item $\gleft(\bm{t})$ as the rectangle $r_i$, $i \in [k]$, satisfying $\xleft(r_i)$ $= \xleft(B_\bm{t})$. In case multiple values in $[k]$ fulfill the condition, let $i$ be the smallest of such values.

@@

In the {\em reporting version} of Problem $\mathscr{A}$, the goal is to output, for each point $p \in P$, all the rectangles $r \in R$ containing $p$; if no such $r$ exists, report nothing for $p$. In Figure~\ref{fig:guard}a, for instance, the output is $\set{(p_2: r_1, r_2), (p_3: r_2)}$. As shown in Appendix~\ref{app:bricks}, the problem can be solved in $O(n \log n + \out)$ time, where $\out$ is the number of pairs $(p, r) \in P \times R$ such that $p \in r$.

@@

\begin{figure*}
    \begin{tabular}{cc}
        \includegraphics[height=30mm]{./artwork/prob-a} &
        \includegraphics[height=27mm]{./artwork/prob-b}  \\[2mm]
        (a) Problem $\mathscr{A}$ &
        \hspace{3mm}
        (b) Problem $\mathscr{B}$ \\[3mm]
    \end{tabular}
    %\vspace{3mm}
    \begin{tabular}{ccc}
        \includegraphics[height=30mm]{./artwork/prob-c} &
        \includegraphics[height=30mm]{./artwork/prob-d} &
        \includegraphics[height=30mm]{./artwork/prob-e} \\[2mm]
        (c) Problem $\mathscr{C}$ &
        (d) Problem $\mathscr{D}$ &
        (e) Problem $\mathscr{E}$
    \end{tabular}

@@

\extraspacing {\bf Problem $\bm{\mathscr{E}}$.} The input involves a set $H$ of horizontal segments and a set $R$ of rectangles. The goal is to report, for each segment $h \in H$, the set $\contain_R(h)$ --- defined in \eqref{eqn:contained} --- where the rectangles are sorted by their right boundaries; if $\contain_R(h)$ is empty, output nothing for $h$. Formally, if $r_1, r_2, ..., r_z$ for some $z \ge 1$ are all the rectangles in $\contain_R(h)$, we output $(h: r_1, r_2, ..., r_z)$, provided that $\xright(r_i) \ge \xright(r_{i-1})$ for each $i \in [2, z]$. Figure~\ref{fig:probs}e gives an example where $H = \set{h_1, h_2, h_3}$ and $R = \set{r_1, r_2, r_3}$; the output is $\set{(h_1: r_2, r_1), (h_2: r_3, r_2)}$. In Appendix~\ref{app:bricks}, we explain how to solve the problem in $O(n \log n + \out)$ time where $n = |H| + |R|$ and $\out$ is the number of pairs $(h, r) \in H \times R$ such that $r$ contains $h$.

@@

\extraspacing {\underline{\em Example 4.1.}} To illustrate our algorithm, we will utilize the running example in Figure~\ref{fig:hv:type1:ex}a, where $k = 4$, and $R_1 = \set{\alpha}$ (the solid rectangle), $R_2 = \set{\beta_1, \beta_2}$ (the dashed rectangles), $H$ $=$ $\set{h_1,$ $h_2,$ $..., h_6}$, and $V = \set{v_1, v_2, ..., v_5}$. The set $\J_1$ contains the following tuples: $(\alpha, \beta_2, h_2, v_3)$, $(\alpha, \beta_2, h_2, v_5)$, $(\alpha, \beta_2, h_5, v_3)$, and $(\alpha,$ $\beta_2, h_5, v_5)$. \done
